Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization

### 2020-08-13 MSc_Dissertation_B156476.tex
Tweaked abstract, GEA, substantial addition to Discussion, going through research questions

\begin{abstract}
Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissue can identify genes whose expression is linked to the pathological process.  In contrast to supervised contrastive approaches (e.g. comparing normal with tumour samples to identify over/under expressed genes), unsupervised approaches work with datasets of mixed and unknown tumour status.   Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.  From these metagenes, sets of genes most associated with component of variation can be identified, and when combined with gene enrichment analysis against Gene Ontology (GO) linked annotations, can provide insights unto the underlying biological processes in play.

@@

Three MF methods are applied and compared: non-negative matrix factorization (NMF), independent component analysis (ICA) and principal component analysis (PCA).  While these are all linear factorization methods, they have very different properties which are explained.  A key problem in applying MF is deciding on the number of components to extract -- the factorization rank.  Too high a rank with result in metagenes which are unstable and fail to generalise to other datasets.  From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by use of bootstrap resampling.  The similarity of the metagenes across the three methods is assessed via the Jaccard index.

An integrative analysis of the GO enrichment results, survival analysis, genomic feature correlations and metagene similarity allows some tentative biological findings to be suggested and corroboration sought in the literature.  The main methodological findings are that MF is effective in identifying and removing technical batch effects, but to beware of instability due to sampling error when applying MF to small datasets.   The question of which of the three methods -- NMF, ICA or PCA -- is most effective for these purposes is considered but not clear conclusion can be drawn.

@@

The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.   
They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.

@@

Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
CD38 is over expressed in myloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.

@@

Dimensionality reduction concerns representing high dimensional data in far fewer dimensions, with minimal loss of information.  This is possible in many cases because dimensions -- or equivalently elements of a feature vector -- are highly correlated.  In the case of gene expression analysis, the dimensionality is of the order 20,000, i.e. the number of transcripts (mRMA transcribed genes) being considered.  However, it is known that sets of genes act in concert, their expression thus being correlated, so reducing the true degrees of freedom in the data.  The challenge is to uncover these true degrees of freedom, which constitute our ``patterns of gene expression''.

The three matrix factorization methods considered in this work are described below.  All are forms of dimensionality reduction, but have differing criteria they aim to optimise.  The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
\subsubsection{Non-negative matrix factorization (NMF)}
\label{sec-nmf}
As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.   This property make the resulting factorization more interpretable since the components are strictly additive.   
NMF has been applied widely in -omics research, originally to gene expression microarrays  and latterly to RNA-seq datasets.  
A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation.
\cite{Brunet2004}.
%
\subsubsection{Independent component analysis (ICA)} 
The goal of ICA is to represent the given matrix with components which are statistically independent of each other.  It was originally proposed to solve the blind source separation problems \cite{Comon1994}. If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.    Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.  See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
\subsubsection{Principal component analysis (PCA)}
PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.  It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can assumed to be so.  PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.

@@

Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.  They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders  and variational autoencoders .  Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.   Their conclusion was that no single method was optimal, but best performance -- judged by capturing the most biological pathway-associated features -- was achieved by taking the extracted latent dimensions of all the methods, an approach they call the ``BioBombe''.

Meng \etal\ \cite{Meng2016} extensively review dimensionality reduction methods applied to multi-omic datasets.   PCA, ICA and NMF are discussed, but the main focus is on less well known algorithms applicable to multiple datasets, including multiple coinertia analysis and consensus PCA.  The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a common and manageable representation space.

@@

\begin{enumerate}
\item Which are the influential genes, or groups of genes, active in HGSOC according to gene expression data?   
\item How do these patterns of gene expression relate to underlying biological processes.  Do these confirm published results?
\item Which MF method is best suited to this analysis and how should they be applied to achieve robust results in the presence of limited dataset size?
\item Do MF methods yield factors which are predictive of patient survival?
\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
\end{enumerate}

@@

\subsubsection{Datasets}
Two gene expression datasets were assailable for this work (n refers to number of patients, g refers to number of genes).
\begin{enumerate}
\item The Cancer Genome Atlas (TCGA) derived, n=274, g=19,601, with metadata on survival
\item Austrailian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features. Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
\end{enumerate}
TODO: \emph{more explanation of the provenance of these datasets}.

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.

@@

\subsubsection{Tools}
\begin{itemize}
\itemsep -0.5em 
\item DELL XPS 13, 16GB RAM, running Linux Mint version v19.
\item Python v3.6.8
\item PyCharm Integrated Development Environment, v2020.1.3
\item Numpy for high performance matrix manipulation, v1.18.15
\item Pandas for data frame handing, v1.0.5
\item Matplotlib and Seaborn for general plotting, v3.2.2 and v0.10.1
\item SciPy for least-square solutions and statistical tests, v1.5.0
\item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
\item GOATOOLS package for gene enrichment analysis against GO, v1.0.6
\item Lifelines package for survival analysis, v0.24.16.
\item Github for source control.
\end{itemize}

@@

An alternative to consider the intersection of the gene sets which each metagene highlights.  In the gene enrichment analysis which follows, candidate genes will be identified for each meatagene.  A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:

@@

\subsection{Gene enrichment analysis}

The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in HGSOC.  
In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.

The essence of gene enrichment analysis is to compare a candidate set of genes with many functional gene sets, asking the question ``are there significantly more (or less, i.e. depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.

@@

Uncorrected p-value threshold was set to 0.01.
Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the FDR threshold being set to 0.01.  This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single gene enrichment analysis run.

@@

The described methodology was implemented in Python.
%except for the KEGG gene enrichment analysis which was implemented in R.
All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  A base class is used to present a uniform interface for the three factorization methods.  Intermediate results are cached with lazy computation for efficient working without ordering dependency.  Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.   The fully automated analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.  The total codebase is approximately 2,900 lines of Python.

@@

Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.  
The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
\caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.  In the case of WGD which has binary values (0, 1), jitter is applied for visual effect only.   Above each plot is shown a correlation coefficient $r$ and associated p-value, highlighted with '***' where $p < 0.01$.  In the case of WGD , the Point-Biserial correlation is used; for all others Pearson's $r$ is used.}
\label{fig-genomic-scatters}
\end{center}
\end{figure}

@@

\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.  The next column gives a terse summary of the gene enrichment analysis results.  The final column lists the metagenes which are apparently similar -- having a Jaccard similarity of $\geq 0.35$.}

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5
, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
\caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
\label{fig-heatmap-BOTH}
\end{center}
\end{figure}

@@

The high level findings of this work are discussed below, under biological and methodological headings.

\subsection{Biological findings}

The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.   Of course Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.  We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated.

We found a weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding.   Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.   One paper suggests a link : ``a comprehensive review of the literature [] reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.   Seek and ye shall find!

Expression signature correlations with mutational load and copy-number variation (CNV) load have not been detected.  These markers of genome instability are known to be very noisy, so if a correlation with expression does exist it would likely be a weak signal, below the detection power of the small (N=80) AOCS dataset.

We found no convincing evidence of a metasample correlation with patient survival.

\subsection{Methodological findings}

@@

We have been able to confirm the observations of \cite{Zhou2018a,Renard2016,Meng2016} that detection and removal of technical  batch effects naturally emerge from MF methods.  That is, the technical and biological variation largely (not completely) separate into different components.  This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as the TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
\caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}
\label{fig-genomic-scatters-BOTH-AOCS}
\end{center}
\end{figure}

>>>>

\begin{abstract}
Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological process.  In contrast to supervised contrastive approaches (e.g. comparing normal with tumour samples to identify over/under expressed genes), unsupervised approaches work with datasets of mixed and unknown tumour status.   Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.  From these metagenes, sets of genes most associated with component of variation can be identified, and when subjected to gene enrichment analysis against the Gene Ontology (GO) can provide insights into the underlying biological processes in play.

@@

Three MF methods are applied and compared: non-negative matrix factorization (NMF), independent component analysis (ICA) and principal component analysis (PCA).  While these are all linear factorization methods, they have very different properties.  A key problem in applying MF is deciding on the number of components to extract -- the factorization rank.  Too high a rank will result in metagenes which are unstable and fail to generalise to other datasets.  From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by  bootstrap resampling then assessing  coherence through k-means clustering and the Silhouette score.  The similarity of the metagenes found by the three methods is assessed via the Jaccard index.

An integrative analysis of the GO enrichment results, survival analysis, genomic feature correlations and metagene similarity allows some tentative biological findings to be suggested and corroboration sought in the literature.  The main methodological findings are that MF is effective in identifying and removing technical batch effects, but to beware of instability due to sampling error when applying MF to small datasets.   The question of which of the three methods -- NMF, ICA or PCA -- is most effective for these purposes is considered but no clear conclusion can be drawn.

@@

The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
Understanding and predicting such resistance is therefore of clinical importance and is addressed by Mairinger \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.   
They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.

@@

Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
CD38 is over expressed in myeloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.

@@

Dimensionality reduction concerns representing high dimensional data in far fewer dimensions, with minimal loss of information.  This is possible in many cases because dimensions -- or equivalently elements of a feature vector -- are highly correlated.  In the case of gene expression analysis, the dimensionality is of the order 20,000, i.e. the number of transcripts (mRNA transcribed genes) being considered.  However, it is known that sets of genes act in concert, their expression thus being correlated, so reducing the true degrees of freedom in the data.  The challenge is to uncover these true degrees of freedom, which constitute our ``patterns of gene expression''.

The three matrix factorization methods considered in this work are described below.  All are forms of dimensionality reduction, but have differing criteria they aim to optimise.  The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
\subsubsection{Non-negative matrix factorization (NMF)}
\label{sec-nmf}
As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.   This property make the resulting factorization more interpretable since the components are strictly additive.   
NMF has been applied widely in -omics research, originally to gene expression microarrays  and latterly to RNA-seq datasets.  
A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation.
\cite{Brunet2004}.
%
\subsubsection{Independent component analysis (ICA)} 
The goal of ICA is to represent the given matrix with components which are statistically independent of each other.  It was originally proposed to solve the blind source separation problems \cite{Comon1994}. If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.    Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.  See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
\subsubsection{Principal component analysis (PCA)}
PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.  It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can be assumed to be so.  PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.

@@

Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.  They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders (DAE)  and variational autoencoders (VAE).  Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.   Their conclusion was that no single method was optimal, but best performance -- judged by capturing the most biological pathway-associated features -- was achieved by taking the extracted latent dimensions of all the methods, an approach they call the ``BioBombe''.

Meng \etal\ \cite{Meng2016} extensively review dimensionality reduction methods applied to multi-omic datasets.   PCA, ICA and NMF are discussed, but the main focus is on less well known algorithms applicable to multiple datasets, including multiple coinertia analysis and consensus PCA.  The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a crommon and manageable representation space.

@@

\begin{enumerate}
\item What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression?  Do these confirm published results?
\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
\item Do MF methods yield factors which are predictive of patient survival?
\item How should MF methods be applied to achieve robust results on datasets of limited size?
\item Do MF detect and remove technical batch effects?
\item Which MF method is best suited to transcripomics analysis?
\end{enumerate}

@@

\subsubsection{Datasets}
Two gene expression datasets were assailable for this work (n refers to number of patients, g refers to number of genes).
\begin{enumerate}
\item The Cancer Genome Atlas (TCGA) derived, n=274, g=19,601, with metadata on survival
\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features. Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
\end{enumerate}
TODO: \emph{more explanation of the provenance of these datasets}.

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis (GEA) against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.

@@

\subsubsection{Tools}
\begin{itemize}
\itemsep -0.5em 
\item DELL XPS 13, 16GB RAM, running Linux Mint version v19.
\item Python v3.6.8
\item PyCharm Integrated Development Environment, v2020.1.3
\item Numpy for high performance matrix manipulation, v1.18.15
\item Pandas for data frame handing, v1.0.5
\item Matplotlib and Seaborn for general plotting, v3.2.2 and v0.10.1
\item SciPy for least-square solutions and statistical tests, v1.5.0
\item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
\item GOATOOLS package for GEA against GO, v1.0.6
\item Lifelines package for survival analysis, v0.24.16.
\item Github for source control.
\end{itemize}

@@

An alternative to consider the intersection of the gene sets which each metagene highlights.  In the GEA which follows, candidate genes will be identified for each meatagene.  A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:

@@

\subsection{Gene enrichment analysis (GEA)}
\label{sec-gea}
The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in HGSOC.  
In order to gain insights into what biological processes are involved, GEA against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.

The essence of GEA is to compare a candidate set of genes with many functional gene sets, asking the question ``are there significantly more (or less, i.e. depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.

@@

Uncorrected p-value threshold was set to 0.01.
Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the FDR threshold being set to 0.01.  This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single GEA run.

@@

The described methodology was implemented in Python.
%except for the KEGG GEA which was implemented in R.
All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  A base class is used to present a uniform interface for the three factorization methods.  Intermediate results are cached with lazy computation for efficient working without ordering dependency.  Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.   The fully automated analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.  The total codebase is approximately 2,900 lines of Python.

@@

GEA against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.  
The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5,draft]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
\caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.  In the case of WGD which has binary values (0, 1), jitter is applied for visual effect only.   Above each plot is shown a correlation coefficient $r$ and associated p-value, highlighted with '***' where $p < 0.01$.  In the case of WGD , the Point-Biserial correlation is used; for all others Pearson's $r$ is used.}
\label{fig-genomic-scatters}
\end{center}
\end{figure}

@@

\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.  The next column gives a terse summary of the GEA results.  The final column lists the metagenes which are apparently similar -- having a Jaccard similarity of $\geq 0.35$.}

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
\caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
\label{fig-heatmap-BOTH}
\end{center}
\end{figure}

@@

Addressing each of our research questions in turn:

\subsubsection*{What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression? }

Biological processes identified by GEA have been set out in section \ref{sec-gea}.   Processes relating to the ECM and extracellular organisation may reflect cell type heterogeneity -- perhaps within the tumour its self, or due to sampling of adjacent normal tissue.   Processes of negative regulation of angiogenesis could reflect tumour growth and therefore be of prognostic value.

These observations are based on GEA on the GO only.  Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways; this was intended but had to be dropped due to shortage of time.

\subsubsection*{How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?}

The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.   Of course, Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.  We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated -- something which warrants further investigation.

Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was foud.   Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.   One paper suggests a link : ``a comprehensive review of the literature reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.   Seek and ye shall find!

Expression signature correlations with mutational load and  CNV load have not been detected.  These markers of genome instability are known to be very noisy, so if a correlation with expression does exist it would likely be a weak signal, below the detection power of the small (N=80) AOCS dataset.

\subsubsection*{Do MF methods yield factors which are predictive of patient survival?}

We found no convincing evidence of this, for an individual factor (metasample).  Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.  The factor with GEA showed indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.   It is possible that a more sophisticated multiple regression model would be successful, but this would risk over-fitting given the limited dataset size.

\subsubsection*{How should MF methods be applied to achieve robust results on datasets of limited size?}

@@

\subsubsection*{Do MF detect and remove technical batch effects?}

Yes. We have been able to confirm the observations of \cite{Zhou2018a,Renard2016,Meng2016} that detection and removal of technical  batch effects naturally emerge from MF methods.  That is, the technical and biological variation largely (not completely) separate into different components.  This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as the TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.

\noindent
\emph{Which MF method is best suited to transcripomics analysis?}

The original focus of this project was ICA and NMF since these are know to be able to disentangle variation in data with complex structure, not multivariate normally distributed.  PCA was added as a baseline dimensionality reduction method, just for comparison.   What would we have been missed if only PCA had been used?  Reviewing the integrative analysis of results in table \ref{tab-integrative-analysis} we see that all three methods picked up on WGD and cellularity.  PCA included the two components having the largest observed effect on overall survival.  NMF and ICA both showed sensitivity to SV load.  NMF further added sensitivity to HRD and identified genes relating to angiogenesis with possible pathological significance.

So it's a mixed picture.  Of course, had more components been selected from each method then likely more would have been found, but at the expense of robustness and thence false discovery, as we have discussed.   It seems we must concur with the conclusion of Way \etal \cite{WayGregory2019} that no one method excels, each brings individual value.   At least these three methods are simple and linear (unlike the VAE for example), and so can be used together easily.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
\caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}
\label{fig-genomic-scatters-BOTH-AOCS}
\end{center}
\end{figure}
